Adding Sensitivity to 21cm Inteferometric Probes of Reionization

### 2016-06-15 Rotation paper/RotationDraft Kopie.tex
updated tex

We define the delay transformed visibility:
\[
\begin{aligned}V(b) & =\int d\nu V_{\nu}(b)\exp\left[-2\pi i\nu\eta\right],\\
 & =\frac{2k_{B}}{\lambda^{2}}\int d\Omega d\nu A(\hat{r},\nu)T(\hat{r},\nu)\exp\left[-2\pi i\left(\frac{\nu}{c}b\cdot\hat{r}+\nu\eta\right)\right]
\end{aligned}
\]

\textbackslash{}The beam reception pattern $A$ is dimensionless,
normalized to 1 at its peak (zenith), and we assume it to be the same
for all baselines. We introduce the Fourier temperature and window
function

@@

\begin{figure}[H]
\includegraphics[scale=0.5]{64layout}

@@

We shall use the 64-antenna PAPER array to demonstrate. The PAPER
64 array is located in the Karoo desert in South Africa (30:43:17.5
S, 21:25:41.8 E). The layout pattern with antenna labels are show
in Fig. 1. I show denote a baseline with the antconfigWe see the antenna
spacing in North-South directions are comparatively close (4m), so
that baselines such as 0\_44 and 0\_7 are very close to equivalent.

\section{Implementation}

An example of rotation tracks is shown in the uv plane below

\begin{figure}[H]
\includegraphics[scale=0.5]{tracks}

The beam squared and its Fourier transform are given as

\begin{figure}[H]
\includegraphics[scale=0.4]{64beam}

\end{document}

>>>>

We define the delay transformed visibility:
\[
\begin{aligned}\end{aligned}
\]

The beam reception pattern $A$ is dimensionless, normalized to 1
at its peak (zenith), and we assume it to be the same for all baselines.
We introduce the Fourier temperature and window function

@@

\begin{figure}[H]
\includegraphics[scale=0.5]{figure_1}

@@

We shall use the 64-antenna PAPER array to demonstrate. The PAPER
64 array is located in the Karoo desert in South Africa (30:43:17.5
S, 21:25:41.8 E). The layout pattern with antenna labels are show
in Fig. 1. I show denote a baseline with the antconfigWe see the antenna
spacing in North-South directions are comparatively close (4m), so
that baselines such as 0\_44 and 0\_7 are very close to equivalent. 
\end{document}

### 2016-06-15 Rotation paper/draft_working.tex
updated tex

%% LyX 2.1.4 created this file.  For more info, see http://www.lyx.org/.
%% Do not edit unless you really know what you are doing.
\documentclass[10pt,english]{article}
\usepackage[T1]{fontenc}
\usepackage[latin9]{inputenc}
\usepackage{geometry}
\geometry{verbose,tmargin=2cm,bmargin=2cm,lmargin=2cm,rmargin=2cm}
\usepackage{color}
\usepackage{babel}
\usepackage{float}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{esint}
\usepackage[unicode=true,
 bookmarks=true,bookmarksnumbered=true,bookmarksopen=false,
 breaklinks=false,pdfborder={0 0 0},backref=false,colorlinks=true]
 {hyperref}
\usepackage{breakurl}

@@

\textbackslash{}The beam reception pattern $A$ is dimensionless,
normalized to 1 at its peak (zenith), and we assume it to be the same
for all baselines. We introduce the Fourier temperature and window
function

\[
\begin{aligned}\tilde{T}(k) & =\int\frac{d^{3}r}{X^{2}Y}T(r)e^{-ikr},\\
W(k) & =\int d^{3}rA(r)e^{-ikr}.
\end{aligned}
\]
By the convolution theorem, we have
\[
\begin{aligned}\tilde{V}(k) & \approx\frac{2k_{B}}{\lambda^{2}}\int\frac{d^{3}r}{X^{2}Y}A(r)T(r)e^{-ikr},\\
 & =\frac{2k_{B}}{\lambda^{2}}\frac{1}{(2\pi)^{3}}\int d^{3}k_{1}W(k_{1})\tilde{T}(k+k_{1}).
\end{aligned}
\]
With implicit bounds of integrals from $-\infty$ to $\infty$, we
have
\begin{equation}
\begin{aligned}\langle\tilde{V}^{*}(k-k_{r})\tilde{V}(k)\rangle & =\left(\frac{2k_{B}}{\lambda^{2}}\right)^{2}\frac{1}{(2\pi)^{6}}\int dk_{1}dk_{2}W^{*}(k_{1})W(k_{2})\langle T^{*}(k+k_{1}-k_{r})T(k+k_{2})\rangle,\\
 & =\left(\frac{2k_{B}}{\lambda^{2}}\right)^{2}\frac{1}{(2\pi)^{6}}\int dk_{1}dk_{2}W^{*}(k_{1})W(k_{2})\delta_{D}(k_{2}-k_{1}-k_{r})\hat{P_{21}}(k+k_{2}),\\
 & \approx\left(\frac{2k_{B}}{\lambda^{2}}\right)^{2}\frac{\hat{P_{21}}(k)}{(2\pi)^{6}}\int dk_{1}W^{*}(k_{1})W(k_{1}+k_{r}),\\
 & =\left(\frac{2k_{B}}{\lambda^{2}}\right)^{2}\frac{\hat{P_{21}}(k)}{(2\pi)^{3}}\int d^{3}r|A(r)|^{2}e^{-ik_{r}r},\\
 & =\left(\frac{2k_{B}}{\lambda^{2}}\right)^{2}\frac{\hat{P_{21}}(k)}{(2\pi)^{3}}\int d^{3}r|a(r)\phi(r_{z})|^{2}e^{-ik_{r}r}
\end{aligned}
\label{eq:(7)}
\end{equation}
where the first to second line follows from assumption of Gaussian
random sky, and the second to third line follows from the assumption
that the 3D power spectrum varies neglegibly over the region of interest
so that $\hat{P_{21}}(k+k_{2})\approx\hat{P_{21}}(k)$. In the last
line we have separated the 2D spatial beam patter from the frequency
bandpass $\phi$. So in conclusion the cross correlation of visibilities
seperated by $\Delta u$ apart in uv space is proportional to the
power spectrum times the Fourier transform of the beam pattern squared.

@@

\section{Implementation}

An example of rotation tracks is shown in the uv plane below

@@

@@

\end{document}

>>>>

%% LyX 2.1.4 created this file.  For more info, see http://www.lyx.org/.
%% Do not edit unless you really know what you are doing.
\documentclass[10pt,english]{article}
\usepackage[T1]{fontenc}
\usepackage[latin9]{inputenc}
\usepackage{geometry}
\geometry{verbose,tmargin=2cm,bmargin=2cm,lmargin=2cm,rmargin=2cm}
\usepackage{color}
\usepackage{babel}
\usepackage{float}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{esint}
\usepackage[unicode=true,
 bookmarks=true,bookmarksnumbered=true,bookmarksopen=false,
 breaklinks=false,pdfborder={0 0 0},backref=false,colorlinks=true]
 {hyperref}

@@

We'd like to relate this to the 3 dimensional power spectrum. We start
by noting that 
\[
\begin{aligned}r & =\frac{c}{H_{0}}\int_{0}^{z}\frac{dz'}{E(z')}-\frac{c(1+z)^{2}}{\nu_{21}H_{0}E(z)}\left(\nu-\nu_{21}\right),\\
 & \equiv D_{c}-Y\Delta\nu,
\end{aligned}
\]
where $\nu_{21}=1420MHz$ is the 21cm transition rest frequency, and
\[
E(z)=\sqrt{\Omega_{m}(1+z)^{3}+\Omega_{\Lambda}}.
\]
So
\[
\nu=\frac{D_{c}-r}{Y}+\nu_{21}.
\]
Thus $d\nu=-dr/Y$ and we can rewrite the delayed transformed visibility
as 
\[
\begin{aligned}V(b) & =\frac{1}{Y}\int_{H}d^{3}rA(r)I(r)\exp\left[2\pi i\frac{r}{Y}\left(\frac{b\cdot\hat{r}}{c}+\eta\right)\right]\exp\left[-2\pi i\left(\frac{b}{c}\cdot\hat{r}+\eta\right)\left(\frac{D_{c}}{Y}+\nu_{21}\right)\right]\end{aligned}
.
\]
$A$ is a patch of shell

The beam reception pattern $A$ is dimensionless, normalized to 1
at its peak (zenith), and we assume it to be the same for all baselines.
We introduce the Fourier temperature and window function

\[
\begin{aligned}\tilde{T}(k) & =\int\frac{d^{3}r}{X^{2}Y}T(r)e^{-ikr},\\
W(k) & =\int d^{3}rA(r)e^{-ikr}.
\end{aligned}
\]
Here 
\[
k=2\pi\frac{b}{cY}.
\]
By the convolution theorem, we have
\[
\begin{aligned}\tilde{V}(k) & =\frac{2k_{B}}{\lambda^{2}}\int\frac{d^{3}r}{Y}A(r)T(r)e^{-ikr},\\
 & =\frac{2k_{B}}{\lambda^{2}}\frac{1}{(2\pi)^{3}}\int d^{3}k_{1}W(k_{1})\tilde{T}(k+k_{1}).
\end{aligned}
\]
With a time offset, the sky has moved from coordinates $r$ to $\Gamma r$.
With implicit bounds of integrals from $-\infty$ to $\infty$, we
have

\[
\begin{aligned}\langle V^{*}(k)V(k')\rangle & =\left(\frac{2k_{B}}{X^{2}Y\lambda^{2}}\right)^{2}\int d^{3}rd^{3}r'\left(\langle T^{*}(r)T(\Gamma r')\rangle\right)A^{*}(r)A(r')\Phi(b,\eta),\\
 & =\left(\frac{2k_{B}}{X^{2}Y\lambda^{2}}\right)^{2}\int d^{3}rd^{3}r'\left(\int\frac{d^{3}\kappa}{(2\pi)^{3}}\frac{d^{3}\kappa'}{(2\pi)^{3}}\langle T^{*}(\kappa)T(\kappa')\rangle e^{-i(\kappa r-\kappa'\Gamma r')}\right)A^{*}(r)A(r')\Phi(b,\eta),\\
 & =\left(\frac{2k_{B}}{X^{2}Y\lambda^{2}}\right)^{2}\int d^{3}rd^{3}r'\left(\int\frac{d^{3}\kappa}{(2\pi)^{3}}P(\kappa)e^{-i\kappa(r-\Gamma r')}\right)A^{*}(r)A(r')\Phi(b,\eta),\\
 & =\left(\frac{2k_{B}}{X^{2}Y\lambda^{2}}\right)^{2}\int d^{3}rd^{3}r'\xi(r-\Gamma r')A^{*}(r)A(r')\Phi(b,\eta),\\
 & \approx\left(\frac{2k_{B}}{X^{2}Y\lambda^{2}}\right)^{2}P(k)\int d^{3}rd^{3}r'\delta_{D}^{(3)}(r-\Gamma r')A^{*}(r)A(r')\exp\left[-i\frac{2\pi}{c}\left(b\cdot\nu\hat{r}-b'\cdot\nu'\hat{r'}\right)\right]\exp\left[-i2\pi\eta\left(\nu\hat{r}-\nu'\hat{r'}\right)\right],\\
 & =\left(\frac{2k_{B}}{X^{2}Y\lambda^{2}}\right)^{2}P(k)\int d^{3}r|A^{*}(r)A(\Gamma r)|\exp\left[-i2\pi\nu_{r}\left\{ \hat{r}\cdot(\frac{b}{c}+\eta c\hat{r})-\hat{\Gamma r}\cdot(\frac{b'}{c}+\eta'\hat{\Gamma r})\right\} \right],
\end{aligned}
\]
where the phase term is 
\[
\Phi(b,\eta)=\text{\ensuremath{\exp\left[-i\frac{2\pi}{c}\left(b\cdot\nu\hat{r}-b'\cdot\nu'\hat{r'}\right)\right]\exp\left[-i2\pi\eta\left(\nu\hat{r}-\nu'\hat{r'}\right)\right]},}
\]
The first to second line follows from assumption of Gaussian random
sky, and the second to third line follows from the assumption that
the 3D power spectrum varies neglegibly over the region of interest
so that $\hat{P_{21}}(k+k_{2})\approx\hat{P_{21}}(k)$. In the last
line we have written $\nu_{r}$ to remind us that $\nu$ and $r$
are related by Eq. 4. Since $\Gamma$ is a sky rotation, it doesn't
affect $\nu$, hence we have taken $\nu_{r}$ outside the parenthesis.
Notice also that we have omitted a phase factor $ $ because it would
drop out in the last step.

Now plugging in result from Eq. 4, we see

\[
\boxed{\langle V^{*}(k)V(k')\rangle=P(k)\left(\frac{2k_{B}}{X^{2}Y\lambda^{2}}\right)^{2}\Psi\int d^{3}r|A^{*}(r)A(\Gamma r)|\exp\left[i\frac{2\pi}{cY}\vec{r}\cdot\left(b-b'\right)\right],}
\]
where $\Psi=\exp\left[-2\pi i\frac{b-b'}{c}\cdot\hat{r}\left(\frac{D_{c}}{Y}+\nu_{21}\right)\right]$
is a phase that's fixed.  So in conclusion the cross correlation of
visibilities seperated by $\Delta u$ apart in uv space is proportional
to the power spectrum times the Fourier transform of the cross multiplied
beam pattern. One can then in principle combine information from different
baseline pairs if one corrects for the phase and normalization.

@@

\section{Technique}

For a given point source in the sky, each baseline at a particular
instant corresponds to a point in UV space. Thus as the earth rotates,
the point source traces out a track in UV plane for a given baseline.
As an example, we show the rotation tracks for a source that passes
through zenith in Fig. \ref{fig:Tracks-of-PAPER64} below for the
baselines of PAPER 64:

@@

\caption{Tracks of PAPER64 for a hypothetical source that passes through zenith.
Color scale simply represents different tracks. Spacing between dots
are characteristic of time resolution of compressed data. As the earth
rotates, tracks are traced out counterclockwise. \label{fig:Tracks-of-PAPER64}}
\end{figure}

@@

In practice, we first implement a search algorithm that grids the
uv plane and search for intersections, then confirm the time delay
by plotting a) cross correlated visibilities from simulated random
sky, and b) theoretical proportion factor $\int d^{3}r|A^{*}(r)A(\Gamma r)|\exp\left[-i\frac{2\pi}{c}\nu\hat{r}\cdot\left(b-b'\right)\right]$
from Eq. 10.

\subsection{Grid search algorithm}

The gridding algorithm sets up grids on the UV-plane, creates a point
source that passes through zenith, and searches for intersections
of the UV tracks. The search compares each track to each other track
to locate the closest approaches. To optimize performance, the algorithm
first sets up coarse grids, locate the cells that contain data points
from more than one tracks, then fine grains the cell to determine
the point of closest approach. Fig {[}insert{]} illustrates this procedure.
This algorithm is mainly used to determine which baselines to cross
correlate, and give a rough estimate of the time offset. We indeed
find that baselines 0\_44 and 0\_7 give the closest approach wiht
a time offset of roughly 0.03 sidereal days. In reality, a wide beam
(e.g. Fig. 2) samples a patch of sky, not just a point source. And
thus the real UV representation of a baseline is the track convolved
with an appropriate point spread function. Accounting for these factors
accurately in this context is difficult. Thus for accurate determination
of the time offset, we resort to two other methods. It is also worth
pointing out that the near-redundant pairs is not only useful at a
particular time corresponding to the crossing. The search algorithm
uses a fiducial point source in the sky to find the offset, and all
time-stamps separated by this offset corresponds to UV-track crossing
of a point in the sky, and thus all data of the two baselines can
be utilized.

\subsection{Gaussian sky simulation}

Having found the pairs of baselines to cross-correlate, and a rough
estimate of the time offset, we make a more accurate determination
of the time offset. Here we take N random realizations of the sky
on a healpix map\footnote{We use functionalities in the python package AIPY for healpix mapping
as well as coordinate transforms. }. For each realization, each pixel is given a Gaussian random value
of brightness temperature. We then rotate the baseline positions by
a simple rotation matrix, recording the visibilities, for each baseline,
as given in Eq. 1 \footnote{As a caveat, there are two obvious ways to achieve the rotation. One
can either fix the sky and rotate the baselines, or the other way
around. We found however, that we must not physically rotate the sky
map, for the numerical roundoffs due to finite resolutions of the
map turns out to be significant. Thus we let the sky, represented
by the healpix map, be fixed, and rotate the baselines. }. The resulting visibilities for the two baselines are then convolved
via the Fourier convolution theorem, to obtain values of the cross
correlation as a function of time-offset. The peak of the curve then
corresponds to maximum redundancy. The accuracy of this result is
limited by (simulated) cosmic variance and finite spacial resolution,
and hence can be beat down by averaging over a large number of universes.
A numerical estimate of the error of the peak height with $N=1$ is
$\lesssim20\%$, and thus with $N=2500,$ we achieve an error of peak
height $<0.5\%$. An example is shown in Fig. {[}insert{]}

\subsection{Theoretical determination}

To test our understanding of the time-offset, and to verify our derivation
in Eq. 10 makes sense, we must check the curve obtained in Fig. against
one from the multiplicative factor in Eq. 10. In particular, we'd
really like to compare the peak location as well as height. For the
latter, the most useful information comes in its comparison with that
of an equivalent baseline. Thus in Fig. {[}insert{]} below we show
on the left comparison of the convolution of an equivalent baseline,
normalized to 1 at the peak, whose location is at 0 offset as expected.
On the right we show comparison of the baseline pair found in the
grid search. The peaks are normalized with the same factor as the
peak height on the left, i.e. the two plots have thus the same scale
on the vertical axis. The height of the plot on the right thus quantifies
the added sensitivity.

\begin{figure}[H]
\includegraphics[scale=0.25]{redun_agree}

\includegraphics[scale=0.3]{simucorroff}

\section{Application to PAPER 64}

\subsection{Expected Sensitivity contribution}

The latest release of PAPER 64 data uses baselines sep-1,1, sep0,1
and sep1,1 {[}cite Ali2015{]}, and achieved a $2\sigma$ upper limit
of $22.4mK^{2}$. There, the three sets of equivalent baselines are
only cross multiplied by itself. Assuming that each baseline delivers
the same quality of data, the relative contribution to sensitivity
can be estimated 
\end{document}
